Adaptive Video Streaming with Optimized Bitstream Extraction and PID-based Quality Control

### 2016-01-12 tmm.tex
Change the second half part according to reply of editing service

Before using the estimation based on LEM, we conduct an experiment to verify its accuracy. Bitstreams in the experiment are encoded using the JSVM encoder, with the GOP structure illustrated in Fig. \ref{fig:GOP_Structure} and the CABAC entropy coding. Each bitstream contains one base layer and one enhancement layer, with QP = 33 and QP = 27, respectively. The enhancement layer is further divided into 2 Medium Grain Scalability (MGS) layers by configuring the MGS vector to separate the transform coefficients into two groups (containing 4 and 12 coefficients respectively). Random combinations of enhancement packets are removed from a bitstream, and the distortion estimated by the linear error model is compared with the distortion actually calculated after real decoding. Decoding is performed by the JSVM decoder, while other operations are performed with MATLAB. One comparison result for the test sequence {\em Foreman} (only the first 33 frames are used) is shown in Fig. \ref{fig:model_verification}. The horizontal axis denotes the frame index, and the vertical axis for MSE measurement. The solid line represents the actual distortion of the frames, and the dot line represents the distortion estimated by the proposed linear error model. It can be seen that the two lines are very close to each other, which demonstrates the accuracy of estimation using the proposed model. Estimation errors for all the eight SVC standard test sequences are listed in Table \ref{tab:estimation-error}, which shows that the average error is only about 5\%.

\begin{figure}[t]
\centering
\includegraphics[width = 0.9\linewidth]{ModelVerification.png}
\caption{Comparison of actual and estimated distortion for sequence {\em Foreman} \label{fig:model_verification}}
\end{figure}

@@

In this subsection, we utilize the linear error model to enable R-D optimized bitstream extraction. Since finding the global optimal solution of the bitstream extraction problem is impractical, our bitstream extraction method uses a greedy algorithm \cite{GreedyAlgo} to get a sub-optimal solution. Note that the greedy algorithm is able to find a local optimum. However, the gap between the local optimum and the global optimum is uncertain. Similar to the greedy solution of the ``0-1 Knapsack Problem", every time a packet is to be discarded, we choose the packet that has the minimal R-D impact. A packet's R-D impact is mathematically measured by:
\begin{equation}
\label{eq:rd_impact}
\Phi = \dfrac{\partial D}{\partial R} \: ,
\end{equation}
where $\partial D$ represents the distortion change brought by discarding this packet, and $\partial R$ is the corresponding rate change. $\partial R$ is simply equivalent to the size of this packet, while $\partial D$ needs to be calculated by subtraction of distortions before and after the packet is discarded.

We simulate a process of discarding all the enhancement packets from a stream, using the above greedy strategy. During this process, the discarding order of a packet is regarded as its priority and recorded. Then the packets' priority values can be stored in the bitstream (either in the priority identifier field of the NALU header or in supplementary enhancement information messages) for real extraction in the future.

For long video sequences, an optimization window containing a certain number of frames is needed, and the priority assignment and bitstream extraction is conducted independently in each window. Suppose the size of the optimization window, i.e., the number of frames involved in the above process, is $K$, and each frame contains $L_Q$ quality layers (one base layer and $L_Q-1$ enhancement layers). Then the total number of enhancement packets to be discarded is $K \cdot (L_Q-1)$, and in every greedy step, there are at most $K$ packets, i.e., the packets currently at the top of every frame (see analysis in \ref{subsec:analysis-extraction}), that can be selected to discard. Given that, the priority assignment process in each optimization window can be described in Algorithm \ref{algo:greedy}, with some terms in it explained below. The initial value $e_{full}$ is the error between the video sequence decoded from full enhancement packets and the original video sequence, as already introduced in (\ref{eq:subset_error}). $\Phi_m$ is defined as:
\begin{equation}
\label{eq:R-D_impact_m}
\Phi_m = \min_{i \in I_{top}} \Phi_i \: ,
\end{equation}
where $I_{top}$ stands for the top-layer packets, those currently able to be discarded, and $\Phi_i$ denotes the RD impact of packet $i$. $\Phi_i$ is calculated based on (\ref{eq:rd_impact}):
\begin{equation}
\label{eq:R-D_impact_i}
\Phi_i = \dfrac{\partial D}{\partial R} = \dfrac{PSNR(e_{seq}) - PSNR(e_{seq} + e_i)}{SIZE(i)} \:,
\end{equation}
where $SIZE(i)$ is the size of packet $i$, and $PSNR(*)$ is the function to obtain PSNR from an error vector. To be clear, PSNR is defined via mean squared error (MSE), i.e., the average of the squares of the elements in the error vector.

\begin{algorithm}
\caption{Greedy-like priority assignment algorithm}
\label{algo:greedy}
\begin{algorithmic}
    \STATE Initialize $e_{seq}$ to $e_{full}$, and initialize $q$ to 0.
    \WHILE{$q < K \cdot (L_Q - 1)$}
	    \FOR{the (at most) $K$ top layer packets}
		    \STATE Find the packet $m$ which has the least RD impact $\Phi_m$
		\ENDFOR
		\STATE Set the priority value of packet $m$ to $q$
		\STATE Remove $m$ from the sequence
		\STATE Update $e_{seq}$ to $e_{seq} + e_m$, and update $q$ to $q+1$
    \ENDWHILE
\end{algorithmic}
\end{algorithm}

It is easy to see that the computational complexity of this priority assignment algorithm is $O(K^2)$. Note that, the computation in this priority assignment process is actually very insignificant compared with the decoding process needed in the error vector acquisition. With a larger optimization window size $K$, we would have more packets to compare and determine their priorities each time, which will potentially result in more optimized extraction. However, doing this will also bring more computation and, more noticeably, increase the memory used to store the error vectors. Besides, if the priority values are to be stored and signalled using the 6-bit priority identifier field of the NALU header \cite{Amonou07}, only a range of 0$\sim$63 is supported, which also limits the value of $K$.

@@

In some scenarios, the priority assignment and bitstream extraction needs to be conducted without access to the original video sequence. This would make the bitstream extraction a more difficult problem, since the distortion is harder to obtain for lacking of reference signal. The ``no-reference bitstream extraction'' problem is actually more frequently encountered in real practice, since it is more common for the streaming service provider to have the compressed bitstream rather than the video source. In this subsection, we consider this no-reference case and modify the proposed algorithm to meet the special characteristic caused by lacking of reference.

Without the original video sequence, we cannot get the error between the video sequence decoded from full enhancement packets and the original sequence, which is $e_{full}$ used in \ref{algo:greedy}. A straightforward idea is to set $e_{full}$ to zero, meaning all distortions are calculated with respect to the fully reconstructed signal, rather than to the original sequence. However, simply doing so causes the algorithm to perform badly in experiments. It is found that, since discarding some enhancement packets will not bring much significant error, the calculated PSNR values are too large to well represent the real distortion. To address this issue, instead of calculating PSNR, we directly use MSE of all pixels as the distortion criteria in the priority assignment algorithm. This proves to be an effective solution, and the extraction performance turns out to be much better.

After each packet is assigned a priority value and optimized extraction is made possible, the video stream now can well adapt to different bitrates. Next, we propose the quality control algorithm based on the PID mechanism, which is responsible for determining the suitable bitrate for transmission according to current network condition.

@@

PID controller \cite{PID}\cite{Astrom02} is a control loop feedback mechanism commonly used in industrial control systems. \textit{Process variable} and \textit{setpoint} are two core concepts in PID control. The process variable stands for the current status of the system. The setpoint is the target for this system to reach. The goal of PID control is to minimize the error between process variable and setpoint. The controller output $u(t)$ is defined as follows:
\begin{equation}
\label{eq:pid-output}
u(t) = {K_p} \cdot e(t) + {K_i} \cdot \int_0^t {e(\tau )d\tau }  + {K_d} \cdot \frac{d}{{dt}}e(t) \: ,
\end{equation}
where $e(t)$ represents the error between process variable and setpoint at time $t$. $K_p$, $K_i$, and $K_d$ here are tuning parameters representing the proportional gain, integral gain and derivative gain, respectively. The result of (\ref{eq:pid-output}) is used for manipulating the system to make process variable change towards setpoint.

As an effective and flexible control technique, PID control has been used to solve all kinds of control problems. In the area of video coding, we notice that PID is utilized to design rate control algorithms (rate control in video coding is mainly to adjust the encoding parameters for every frame, so that the bits are appropriately allocated and desired bitrate is achieved) \cite{Wong04}\cite{Yang10}. We also notice that some researchers have proposed theoretical model or middle-ware framework for QoS adaptation \cite{Li98}\cite{Li99}, with PID as the basic control scheme. However, we have not found any paper that integrates PID method into the quality control of video streaming. While rate control of video coding and quality control of video streaming are two different problems, they do share some common properties, such as the buffer mechanism and the requirement for smoothness. That inspires us to design a PID-based quality control algorithm and utilize it to deliver better video quality to the user.

In this section, we first introduce the \textit{quality level} concept to discretize the bitrate and quality of scalable video. Then we define the process variable used in our PID-based quality control algorithm. After that, the algorithm is formulated and explained in detail, followed by the brief discussion of parameter selection.

@@

The video quality is a continuous value, and it is impractical to control it at an arbitrary precision. Therefore we propose a \textit{quality level} concept as the discrete description of the video quality. When the video quality needs to be adjusted, we simply increase or decrease the quality level. Each quality level corresponds to a certain bitrate. A higher quality level corresponds to a higher bitrate and vice versa. The number of quality levels and how they are mapped to sub-streams with different bitrates depends on actual requirement or implementation. The more quality levels are defined, the more precisely the control algorithm can adjust bitrate to follow the bandwidth change. As pointed out in section \ref{sec:intro}, simulcast streaming systems can only have limited quality levels while SVC streaming systems can define quality levels at a much finer granularity.

@@

In DSS, two timelines are important. Push timeline indicates the time of pushing video data packets to streaming buffer, and play timeline indicates the time these data packet should be played, i.e., the presentation time stamp (often seen as PTS) of each packet. Their relationship is shown in Fig. \ref{fig:intervals}. The program will check the timelines periodically, and, in the default quality control algorithm of DSS, use the packet delays to decide whether quality change is needed. The two intervals shown in Fig. \ref{fig:intervals}, i.e., check interval and actual interval, are worth noticing and may provide useful information. In the ideal situation, during each check interval, the server pushes an amount of data that will actually play for a period of time equal to that check interval, i.e., \textit{actual\_interval} = \textit{check\_interval}. If \textit{actual\_interval} is smaller than \textit{check\_interval}, it means that less data is pushed to the client, and the pushing is slowed down because bandwidth is relatively low at this time. On the other hand, if \textit{actual\_interval} is larger than \textit{check\_interval}, it indicates the network condition is good and data is being pushed quickly to the client. The ratio of these two intervals, i.e., \textit{actual\_check interval ratio}, is in direct proportion to the ratio of the bitrate suitable for the network condition and the bitrate currently being used.

The \textit{actual\_check interval ratio} is suitable to use in quality control because: 1) this ratio directly reflects the mismatch of current pushing bitrate and the bandwidth, and by adjusting bitrate according to it, this ratio can be used to accurately change video quality; 2) the desired value of this ratio is 1.0 by definition so we don't need to find the optimized setpoint by mathematical methods or experiments; and 3) it is easy to compute and implement in practical systems.

\begin{figure}[t]
\centering
\includegraphics[width = 0.9\linewidth]{Intervals.png}
\caption{Two timelines in Darwin Streaming Sever (DSS) \label{fig:intervals}}
\end{figure}

@@

As is introduced above, the \textit{actual\_check interval ratio} serves as the process variable in this PID-based quality control algorithm. In normal PID control, the controller output is calculated using the proportion, integral and differential of the error between the process variable and the setpoint, as shown in (\ref{eq:pid-output}). Considering that the chosen process variable is a ratio and the corresponding setpoint is 1, we make a little modification to the general PID controller and obtain a ratio-based model. In this model, the ratio is directly used to calculate the controller output, and all subtraction operations in original PID equations have been properly changed to division operations to keep the correctness, as follows:

@@

In (\ref{eq:ut}), $K_p$, $K_i$, and $K_d$ indicate the tuning parameters for proportional, integral and derivative parts respectively, and $E_p^t$, $E_i^t$, $E_d^t$ stand for the corresponding observed value for each part, which are defined in (\ref{eq:ep})-(\ref{eq:long-check}); the controller output $u_t$ is utilized to calculate a suitable bitrate for transmission and further to determine which quality level should be selected (see Algorithm \ref{algo:control} for more details).

In (\ref{eq:long-actual}) and (\ref{eq:long-check}), $T_l$ stands for the last quality change time. Theoretically, the integral part should be recorded and calculated since the system starts. In practice, however, bandwidth may change significantly during the transmission. If the integral part is recorded for too long time, it may fail to reflect the current network condition and thus slow down the reaction for better quality level. A compromise here is to choose a relatively long time, i.e., the time from last quality change till now, for the integral part calculation.

This ratio-based model is more simple and effective for the chosen process variable and setpoint. Moreover, in this model the controller output has intuitive meaning when the three parameters are normalized. It is easy to know that $E_p^t$, $E_i^t$, $E_d^t$ are all ratios with a value more or less than 1, and the controller output $u_t$ can be seen as a weighted average of the three ratios, with the three parameters as weights. Therefore the controller output is also a ratio value around 1, which reflects the relationship between the bitrate suitable for the network condition and the bitrate currently being used. In other words, if the output is larger than 1, then the controller has determined the bitrate should be increased accordingly, and vice versa. Use it as a multiplier to manipulate the bitrate currently being used, and we would be able to control the streaming system to match the network condition.

The whole process of the proposed algorithm is described in Algorithm \ref{algo:control}. Compared to other scheduling strategies, this quality control algorithm has several advantages. First, this algorithm does not just rely on current packet delay or information at a fixed time point, therefore quality change may occur at any proper time. Second, since the model has taken prediction into account, it is able to decide whether the bandwidth will be high or low enough for the quality level to change in the near future, thus to avoid unnecessary quality change. Third, in this algorithm it's not obliged to change one quality level at a time. This enables fast quality drop to avoid a latent playback pause when bandwidth suddenly decreases, and also fast quality rise to provide better video as soon as possible when system starts, resulting in a seamless watching experience where the user can enjoy the video without stopping or waiting.

\begin{algorithm}
\caption{PID-based quality control algorithm}
\label{algo:control}
\begin{algorithmic}
    \STATE before packet is sent
    \STATE check\_interval = curr\_time -- last\_check\_time
    \STATE actual\_interval = curr\_media\_time -- last\_check\_media\_time
    \STATE long\_check\_interval += check\_interval
    \STATE long\_actual\_interval += actual\_interval
    \STATE calculate $E_p^t$, $E_i^t$, $E_d^t$ according to Eq. (\ref{eq:ep}) - (\ref{eq:ed})
    \STATE output = ${K_p} \cdot E_p^t + {K_i} \cdot E_i^t + {K_d} \cdot E_d^t$
    \STATE new\_bitrate = output * curr\_bitrate
    \STATE new\_level = get\_level(new\_bitrate, bitrate\_of\_level[])
    \IF {quality level changed}
    	\STATE long\_check\_interval = 0
    	\STATE long\_actual\_interval = 0
    	\STATE curr\_bitrate = bitrate\_of\_level[new\_level]
    \ENDIF
    \STATE send packet according to new quality level
\end{algorithmic}
\end{algorithm}

@@

As with other PID control systems in the industry, the selection of the three parameters is implementation-dependent and most of the time manual tuning with trial and error is required to achieve optimal results. Each parameter has impacts on different aspects of the system, and knowing them is useful in the practice of parameter selection. Generally, the proportional part in PID controller affects the system's stability. If the proportional gain $K_p$ is too high, the system can become unstable. However, a low $K_p$ would result in long rising time that makes the system response slowly. In addition to eliminating steady-state error, the integral term also accelerates the movement of the process towards setpoint, which means increasing $K_i$ can reduce the rising time. However, since the integral part accumulates errors from the past, it can cause the present value to overshoot the setpoint value, and too much $K_i$ will lead to instability as well. These can all be compensated by the derivative part, which predicts system behaviour and has the effect of reducing the overshoot, increasing the stability, and improving the settling time (time it takes to reach setpoint after a disturbance). The big problem with the derivative part is that it is very sensitive to noise, therefore industrial practices indicate that the derivative gain $K_d$ should usually stay low.

For the proposed quality control algorithm, Ziegler--Nichols method \cite{Ziegler42} is used to select the parameters for better performance. This method is performed by setting integral and derivative parameters to zero and then tuning $K_p$ to a value $K_u$ until output begins to oscillate with constant amplitude. Assuming the oscillation period is $T_u$, we set $K_p = 0.6K_u$, $K_i = 2K_u/T_u$, and $K_d = K_u \cdot T_u/8$. $K_p$, $K_i$, and $K_d$ are then normalized to get the final parameters. In the following experiments, parameters are chosen as $K_p = 0.22$, $K_i = 0.73$, $K_d = 0.05$ by this method. With these parameters, the PID-controlled streaming system is stable, and also quickly begins to follow the bandwidth change, as shown by the experimental results in next section.

We choose to use the standard Ziegler--Nichols method because it already takes each parameter's impacts into consideration and produces acceptably good results. Adopting more sophisticated or automated techniques to further optimize the parameter selection and controller performance will be our future work. The robustness of the controller will also be improved by adding a mechanism that dynamically adjusts the controller parameters based on environment and performance change. Note that, in the video quality control algorithm only the bitrate and the bandwidth are relevant, not the video content. Therefore once the parameters are chosen they should apply well to different video streams.

@@

To implement our proposed adaptive streaming system requires the combination of the optimized bitstream extraction and the PID-based quality control algorithm. The latter determines the current bitrate to follow the bandwidth change, and the former will ensure the sub-stream with the best possible quality is obtained and sent under that bitrate constraint. They work together to improve the user's experience of watching online videos. With DSS as the implementation basis, we have integrated the proposed algorithms into it for the purpose of both research and application.

In this section, we first present separate experiments to show the advantage of each proposed solution, then give the result when they are combined together.

@@

\subsection{Result of the Combined System}
\label{subsec:exp-combined}

>>>>

Before using the estimation based on LEM, we conduct an experiment to verify its accuracy. Bitstreams in the experiment are encoded using the JSVM encoder, with the GOP structure illustrated in Fig. \ref{fig:GOP_Structure} and the CABAC entropy coding. Each bitstream contains one base layer and one enhancement layer, with QP = 33 and QP = 27, respectively. The enhancement layer is further divided into 2 Medium Grain Scalability (MGS) layers by configuring the MGS vector to separate the transform coefficients into two groups (containing 4 and 12 coefficients respectively). Random combinations of enhancement packets are removed from a bitstream, and the distortion estimated by the linear error model is compared with the distortion actually calculated after real decoding. Decoding is performed by the JSVM decoder, while other operations are performed with MATLAB. One comparison result for the test sequence {\em Foreman} (only the first 33 frames are used) is shown in Fig. \ref{fig:model_verification}. The horizontal axis denotes the frame index, and the vertical axis denotes the MSE measurement. The solid line represents the actual distortion of the frames, and the dotted line represents the distortion estimated by the proposed linear error model. It can be seen that the two lines are very close to each other, which demonstrates the accuracy of estimation using the proposed model. Estimation errors for all the eight SVC standard test sequences are listed in Table \ref{tab:estimation-error}, which shows that the average error is only about 5\%.

\begin{figure}[t]
\centering
\includegraphics[width = 0.9\linewidth]{ModelVerification.png}
\caption{Comparison of the actual and the estimated distortion for sequence {\em Foreman} \label{fig:model_verification}}
\end{figure}

@@

In this subsection, we utilize the linear error model to enable R-D optimized bitstream extraction. Because finding the global optimal solution of the bitstream extraction problem is impractical, our bitstream extraction method uses a greedy algorithm \cite{GreedyAlgo} to obtain a sub-optimal solution. Note that the greedy algorithm is able to find a local optimum. However, the gap between the local optimum and the global optimum is uncertain. Similar to the greedy solution of the ``0-1 Knapsack Problem", every time a packet is to be discarded, we choose the packet that has the minimal R-D impact. A packet's R-D impact is mathematically measured by:
\begin{equation}
\label{eq:rd_impact}
\Phi = \dfrac{\partial D}{\partial R} \: ,
\end{equation}
where $\partial D$ represents the distortion change brought by discarding this packet, and $\partial R$ is the corresponding rate change. $\partial R$ is simply equivalent to the size of this packet, while $\partial D$ needs to be calculated by subtracting the distortions before and after the packet is discarded.

We simulate the process of discarding all the enhancement packets from a stream, using the above greedy strategy. During this process, the discarding order of a packet is regarded as its priority and recorded. Then, the packet priority values can be stored in the bitstream (either in the priority identifier field of the NALU header or in supplementary enhancement information messages) for real extraction in the future.

For long video sequences, an optimization window containing a certain number of frames is required, and the priority assignment and bitstream extraction are conducted independently in each window. Suppose the size of the optimization window, i.e., the number of frames involved in the above process, is $K$, and each frame contains $L_Q$ quality layers (one base layer and $L_Q-1$ enhancement layers). Then, the total number of enhancement packets to be discarded is $K \cdot (L_Q-1)$, and in every greedy step, there are at most $K$ packets, i.e., the packets currently at the top of every frame (see analysis in \ref{subsec:analysis-extraction}), that can be selected for discarding. Given that, the priority assignment process in each optimization window can be described in Algorithm \ref{algo:greedy}, with some terms in it explained below. The initial value $e_{full}$ is the error between the video sequence decoded from full enhancement packets and the original video sequence, as already introduced in (\ref{eq:subset_error}). $\Phi_m$ is defined as:
\begin{equation}
\label{eq:R-D_impact_m}
\Phi_m = \min_{i \in I_{top}} \Phi_i \: ,
\end{equation}
where $I_{top}$ stands for the top-layer packets, those currently able to be discarded, and $\Phi_i$ denotes the RD impact of packet $i$. $\Phi_i$ is calculated based on (\ref{eq:rd_impact}):
\begin{equation}
\label{eq:R-D_impact_i}
\Phi_i = \dfrac{\partial D}{\partial R} = \dfrac{PSNR(e_{seq}) - PSNR(e_{seq} + e_i)}{SIZE(i)} \:,
\end{equation}
where $SIZE(i)$ is the size of packet $i$, and $PSNR(*)$ is the function for obtaining PSNR from an error vector. For clarification, PSNR is defined using the mean squared error (MSE), i.e., the average of the squares of the elements in the error vector.

\begin{algorithm}
\caption{Greedy-like priority assignment algorithm}
\label{algo:greedy}
\begin{algorithmic}
    \STATE Initialize $e_{seq}$ to $e_{full}$, and initialize $q$ to 0.
    \WHILE{$q < K \cdot (L_Q - 1)$}
	    \FOR{the (at most) $K$ top layer packets}
		    \STATE Find the packet $m$ that has the least RD impact $\Phi_m$
		\ENDFOR
		\STATE Set the priority value of packet $m$ to $q$
		\STATE Remove $m$ from the sequence
		\STATE Update $e_{seq}$ to $e_{seq} + e_m$, and update $q$ to $q+1$
    \ENDWHILE
\end{algorithmic}
\end{algorithm}

It is easy to see that the computational complexity of this priority assignment algorithm is $O(K^2)$. Note that, the computation in this priority assignment process is actually very insignificant compared with the decoding process required for the error vector acquisition. With a larger optimization window size $K$, we would have more packets to compare and determine their priorities each time, which will potentially result in more optimized extraction. However, doing this will also require more computation and, more noticeably, increase the memory used to store the error vectors. Besides, if the priority values are to be stored and signaled using the 6-bit priority identifier field of the NALU header \cite{Amonou07}, only a range of 0$\sim$63 is supported, which also limits the value of $K$.

@@

In some scenarios, the priority assignment and bitstream extraction need to be conducted without access to the original video sequence. This would make the bitstream extraction a more difficult problem, because the distortion is harder to obtain as the reference signal is not available. This ``no-reference bitstream extraction'' problem is actually more frequently encountered in real practice, because it is more common for the streaming service provider to have a compressed bitstream rather than the video source. In this subsection, we consider this no-reference case and modify the proposed algorithm to meet the special characteristics caused by lacking of reference.

Without the original video sequence, we cannot obtain the error between the video sequence decoded from full enhancement packets and the original sequence, which is $e_{full}$ used in \ref{algo:greedy}. A straightforward idea is setting $e_{full}$ to zero, which means that all distortions are calculated with respect to the fully reconstructed signal, rather than the original sequence. However, simply doing so causes the algorithm to perform badly in experiments. It is found that, as discarding some enhancement packets will not result in much significant error, the calculated PSNR values are too large to properly represent the real distortion. To address this issue, instead of calculating the PSNR, we directly use the MSE of all pixels as the distortion criteria in the priority assignment algorithm. This proves to be an effective solution, and the extraction performance turns out to be improved.

After each packet is assigned a priority value and an optimized extraction is made possible, the video stream can now adapt to different bitrates well. Next, we propose a quality control algorithm based on the PID mechanism, which is responsible for determining the suitable bitrate for transmission according to the current network conditions.

@@

The PID controller \cite{PID}\cite{Astrom02} is a control loop feedback mechanism commonly used in industrial control systems. \textit{Process variable} and \textit{setpoint} are two core concepts in PID control. The process variable stands for the current status of the system. The setpoint is the target for this system to reach. The goal of PID control is to minimize the error between the process variable and the setpoint. The controller output $u(t)$ is defined as follows:
\begin{equation}
\label{eq:pid-output}
u(t) = {K_p} \cdot e(t) + {K_i} \cdot \int_0^t {e(\tau )d\tau }  + {K_d} \cdot \frac{d}{{dt}}e(t) \: ,
\end{equation}
where $e(t)$ represents the error between the process variable and the setpoint at time $t$. $K_p$, $K_i$, and $K_d$ here are tuning parameters that represent the proportional gain, integral gain and derivative gain, respectively. The result of (\ref{eq:pid-output}) is used for manipulating the system to make the process variable change towards the setpoint.

As an effective and flexible control technique, PID control has been used to solve many kinds of control problems. In the area of video coding, we notice that PID is utilized to design rate control algorithms (rate control in video coding mainly involves adjusting the encoding parameters for every frame, so that the bits are appropriately allocated and the desired bitrate is achieved) \cite{Wong04}\cite{Yang10}. We also notice that some researchers have proposed a theoretical model or middle-ware framework for QoS adaptation \cite{Li98}\cite{Li99}, with PID as the basic control scheme. However, we have not found any paper that integrates the PID method into the quality control of video streaming. While rate control of video coding and quality control of video streaming are two different problems, they do share some common properties, such as the buffer mechanism and the requirement for smoothness. We are therefore inspired to design a PID-based quality control algorithm and utilize it to deliver better video quality to the user.

In this section, we first introduce the \textit{quality level} concept to discretize the bitrate and quality of a scalable video. Then we define the process variable used in our PID-based quality control algorithm. After that, the algorithm is formulated and explained in detail, followed by a brief discussion on parameter selection.

@@

The video quality is a continuous value, and it is impractical to control it at an arbitrary precision. Therefore, we propose a \textit{quality level} concept for a discrete description of the video quality. When the video quality is to be adjusted, we simply increase or decrease the quality level. Each quality level corresponds to a certain bitrate. A higher quality level corresponds to a higher bitrate, and vice versa. The number of quality levels and how they are mapped to sub-streams with different bitrates depend on the actual requirements or implementation. The more quality levels are defined, the more precisely the control algorithm can adjust bitrate to follow the bandwidth change. As pointed out in section \ref{sec:intro}, simulcast streaming systems can only have limited quality levels while SVC streaming systems can define quality levels at a much finer granularity.

@@

In DSS, two timelines are important. Push timeline indicates the time of pushing video data packets to the streaming buffer, and play timeline indicates the time these data packet should be played, i.e., the presentation time stamp (often seen as PTS) of each packet. Their relationship is shown in Fig. \ref{fig:intervals}. The program will check the timelines periodically, and, in the default quality control algorithm of DSS, use the packet delays to decide whether quality change is needed. The two intervals shown in Fig. \ref{fig:intervals}, i.e., check interval and actual interval, are worth noticing and may provide useful information. In the ideal situation, during each check interval, the server pushes an amount of data that will actually play for a period of time equal to the check interval, i.e., \textit{actual\_interval} = \textit{check\_interval}. If \textit{actual\_interval} is smaller than \textit{check\_interval}, it means that less data is pushed to the client, and the pushing is slowed down because bandwidth is relatively low at this time. On the other hand, if \textit{actual\_interval} is larger than \textit{check\_interval}, it indicates that the network conditions are good and the data are being pushed quickly to the client. The ratio of these two intervals, i.e., \textit{actual\_check interval ratio}, is in direct proportion to the ratio of the bitrate suitable for the network conditions and the bitrate currently being used.

The \textit{actual\_check interval ratio} is suitable for use in quality control because: 1) this ratio directly reflects the mismatch of the current pushing bitrate and the bandwidth, and by adjusting the bitrate according to it, this ratio can be used to accurately change the video quality; 2) the desired value of this ratio is 1.0 by definition, therefore, we do not need to determine the optimized setpoint using mathematical methods or experiments; and 3) it is easy to compute and implement in practical systems.

\begin{figure}[t]
\centering
\includegraphics[width = 0.9\linewidth]{Intervals.png}
\caption{Two timelines in the Darwin Streaming Sever (DSS) \label{fig:intervals}}
\end{figure}

@@

As is introduced above, the \textit{actual\_check interval ratio} serves as the process variable in this PID-based quality control algorithm. In normal PID control, the controller output is calculated using the proportion, integral and differential of the error between the process variable and the setpoint, as shown in (\ref{eq:pid-output}). Considering that the chosen process variable is a ratio and the corresponding setpoint is 1, we make a little modification to the general PID controller and obtain a ratio-based model. In this model, the ratio is directly used to calculate the controller output, and all subtraction operations in the original PID equations have been properly changed to division operations for correctness, as follows:

@@

In (\ref{eq:ut}), $K_p$, $K_i$, and $K_d$ indicate the tuning parameter for the proportional, integral and derivative part, respectively, and $E_p^t$, $E_i^t$, $E_d^t$ stand for the corresponding observed values for each part, which are defined in (\ref{eq:ep})-(\ref{eq:long-check}); the controller output $u_t$ is utilized to calculate a suitable bitrate for transmission and further to determine which quality level should be selected (see Algorithm \ref{algo:control} for more details).

In (\ref{eq:long-actual}) and (\ref{eq:long-check}), $T_l$ stands for the last quality change time. Theoretically, the integral part should be recorded and calculated since the system starts. In practice, however, the bandwidth may change significantly during the transmission. If the integral part is recorded for too long time, it may fail to reflect the current network conditions and thus slow down the reaction for better quality levels. A compromise here is to choose a relatively long time, i.e., the time from last quality change till now, for the calculation of the integral part.

This ratio-based model is more simple and effective for the chosen process variable and setpoint. Moreover, in this model, the controller output has intuitive meaning when the three parameters are normalized. It is easy to see that $E_p^t$, $E_i^t$, $E_d^t$ are all ratios with a value around 1, and the controller output $u_t$ can be seen as a weighted average of the three ratios, with the three parameters as weights. Therefore, the controller output is also a ratio with a value around 1, which reflects the relationship between the bitrate suitable for the network conditions and the bitrate currently being used. In other words, if the output is larger than 1, then the controller has determined that the bitrate should be increased accordingly, and vice versa. Use it as a multiplier to manipulate the bitrate currently being used, and we would be able to control the streaming system to match the network conditions.

The proposed algorithm is described in Algorithm \ref{algo:control}. Compared to other scheduling strategies, this quality control algorithm has several advantages. First, this algorithm does not just rely on the current packet delay or the information at a fixed time point; therefore, a quality change may occur at any proper time. Second, because the model takes prediction into account, it is able to decide whether the bandwidth will be high or low enough for the quality level to change in the near future, so as to avoid unnecessary quality changes. Third, in this algorithm it's not obliged to change one quality level at a time. This enables a fast quality drop to avoid latent playback pause when bandwidth suddenly decreases, as well as a fast quality rise to provide better video as soon as possible when system starts, resulting in a seamless watching experience, where the user can enjoy the video without stopping or waiting.

\begin{algorithm}
\caption{PID-based quality control algorithm}
\label{algo:control}
\begin{algorithmic}
    \STATE before packet is sent
    \STATE check\_interval = curr\_time -- last\_check\_time
    \STATE actual\_interval = curr\_media\_time -- last\_check\_media\_time
    \STATE long\_check\_interval += check\_interval
    \STATE long\_actual\_interval += actual\_interval
    \STATE calculate $E_p^t$, $E_i^t$, $E_d^t$ according to Eq. (\ref{eq:ep}) - (\ref{eq:ed})
    \STATE output = ${K_p} \cdot E_p^t + {K_i} \cdot E_i^t + {K_d} \cdot E_d^t$
    \STATE new\_bitrate = output * curr\_bitrate
    \STATE new\_level = get\_level(new\_bitrate, bitrate\_of\_level[])
    \IF {quality level changed}
    	\STATE long\_check\_interval = 0
    	\STATE long\_actual\_interval = 0
    	\STATE curr\_bitrate = bitrate\_of\_level[new\_level]
    \ENDIF
    \STATE send packet according to the new quality level
\end{algorithmic}
\end{algorithm}

@@

As with other PID control systems in the industry, the selection of the three parameters is implementation-dependent, and most of the time, manual tuning with trial and error is required to achieve optimal results. Each parameter has impacts on different aspects of the system, and knowing these impacts is useful for parameter selection. Generally, the proportional part in the PID controller affects the system's stability. If the proportional gain $K_p$ is too high, the system can become unstable. However, a low $K_p$ would result in a long rising time that makes the system response slow. In addition to eliminating the steady-state error, the integral term also accelerates the movement of the process value towards the setpoint, which means that increasing $K_i$ can reduce the rising time. However, because the integral part accumulates errors from the past, it can cause the present value to overshoot the setpoint value, and a too high $K_i$ will lead to instability as well. These can all be compensated by the derivative part, which predicts the system behavior and has the effect of reducing the overshoot, increasing the stability, and improving the settling time (the time taken to reach setpoint after a disturbance). The issue with the derivative part is that it is very sensitive to noise; therefore, industrial practices indicate that the derivative gain $K_d$ should usually stay low.

For the proposed quality control algorithm, the Ziegler--Nichols method \cite{Ziegler42} is used to select the parameters for better performance. This method is performed by setting the integral and derivative parameters to zero and then tuning $K_p$ to the value $K_u$ until the output begins to oscillate with constant amplitude. Assuming that the oscillation period is $T_u$, we set $K_p = 0.6K_u$, $K_i = 2K_u/T_u$, and $K_d = K_u \cdot T_u/8$. $K_p$, $K_i$, and $K_d$ are then normalized to obtain the final parameters. In the following experiments, the parameters are chosen as $K_p = 0.22$, $K_i = 0.73$, $K_d = 0.05$ using this method. With these parameters, the PID-controlled streaming system is stable, and also quickly begins to follow the bandwidth change, as shown by the experimental results in the next section.

We choose to use the standard Ziegler--Nichols method because it already considers each parameter's impacts and produces acceptably good results. Adopting more sophisticated or automated techniques to further optimize the parameter selection and controller performance will be our future work. The robustness of the controller can also be improved by adding a mechanism that dynamically adjusts the controller parameters based on environment and performance change. Note that in the video quality control algorithm, only the bitrate and the bandwidth are relevant, not the video content. Therefore, once the parameters are chosen, they should apply well to different video streams.

@@

Implementing our proposed adaptive streaming system requires the combination of the optimized bitstream extraction and the PID-based quality control algorithm. The latter determines the current bitrate to follow the bandwidth change, and the former will ensure that the sub-stream with the best possible quality is obtained and sent under that bitrate constraint. They work together to improve the user's experience of watching online videos. With DSS as the implementation basis, we have integrated the proposed algorithms into it for the purpose of both research and applications.

In this section, we first present separate experiments to show the advantage of each proposed solution, and then present the result when they are combined together.

@@

\subsection{Results of the Combined System}
\label{subsec:exp-combined}
